\texttt{MSnbase}, efficient R-based processing and visualisation of raw mass spectrometry data

### 2020-04-27 msnbase2.Rnw
first draft ready

%% \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
%% \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
%% \newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
%% \newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
%% \newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
%% \newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
%% \newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
%% \newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
%% \newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
%% \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
%% \let\hlipl\hlkwb

%% \usepackage{framed}
%% \makeatletter
%% \newenvironment{kframe}{%
%%  \def\at@end@of@kframe{}%
%%  \ifinner\ifhmode%
%%   \def\at@end@of@kframe{\end{minipage}}%
%%   \begin{minipage}{\columnwidth}%
%%  \fi\fi%
%%  \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
%%  \colorbox{shadecolor}{##1}\hskip-\fboxsep
%%      % There is no \\@totalrightmargin, so:
%%      \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
%%  \MakeFramed {\advance\hsize-\width
%%    \@totalleftmargin\z@ \linewidth\hsize
%%    \@setminipage}}%
%%  {\par\unskip\endMakeFramed%
%%  \at@end@of@kframe}
%% \makeatother

%% \usepackage{xcolor}

%% \definecolor{shadecolor}{rgb}{.97, .97, .97}
%% \definecolor{messagecolor}{rgb}{0, 0, 0}
%% \definecolor{warningcolor}{rgb}{1, 0, 1}
%% \definecolor{errorcolor}{rgb}{1, 0, 0}
%% \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

@@

\title[MSnbase version 2]
  {\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data}

@@

Mass spectrometry is a powerful technology to assays chemical and
biological samples. It is used routinely, with well characterised
protocol, as well a development platform, to improve on existing
methods and devise new ones to analyse ever more complex sample in
greater details. The complexity and diversity of mass spectrometry
yields data that is itself complex and often times of considerable
size, that requires non trivial processing before producing
interpretable results. This is particularly relevant, and can
constitue a significant challenge for method developers that, in
addition to the development of sample processing and mass spectrometry
methods, need to process and analyse these new data to demonstrate the
improvement in their technical and analytical work.

There exists a very diverse catalogue of software tools to explore,
process and interpret mass spectrometry data. These range from low
level software libraries such as vendor libraries, jmzML (ref),
proteowizard (ref), ... that are aimed at programmers to develop new
applications, to user-oriented applications, such as
ProteomeDiscoverer, MaxQuant, ... that provide a limited and fixed set
of functionality. The former are used through application programming
interfaces exclusively, while the latter generally featuring graphical
user interfaces (GUI).

TODO: Give examples of libraries re-used in user/gui focused
application...

In this software note, we present version 2 of the \texttt{MSnbase}
\cite{Gatto:2012} software package, available from the
Bioconductor~\cite{Huber:2015} project since 2011. \texttt{MSnbase}
offers a platform that lies between low level libraries and end-use
software. It provides a flexible command line environment for
metabolomics and proteomics mass spectrometry-based application, that
allows a detailed step-by-step processing, analysis and exploration of
the data and development of novel computational mass spectrometry
methods.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a table with
variables along the columns and each row associated to a
spectrum. Among the metadata available for each spectrum, we there is
its MS level, acquisition number, retention time, precursor m/z and
intensity (for MS level 2 and above), and many more. \texttt{MSnbase}
provides a rich interface to manipulate such objects. The code chunk
below illustrates such an object as displayed in the R console and an
enumeration of the metadata.

@@

We will also illustrate how \texttt{MSnbase} makes use of the
forward-pipe operator (\texttt{\%>\%}) defined in the
\texttt{magrittr} package. This operator has proved useful to develop
non-trivial pipelines by combining individual functions into easily
readable operations.

@@

\item Identify and filter the groups of spectra that represent
  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)).. This can be
  done using the `filterString` metadata variable that identifies
  BoxCar spectra by their adjacent M/Z segments with the
  \texttt{bc\_groups()} function and filtering relevant spectra with
  the \texttt{filterBoxCar()}.

@@

The R environment is well known for the quality of its visualiation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data for the serine and the same data after
smoothing, centroiding and M/Z refinement. Detailed execution and
description of these operations can be found in the \emph{MSnbase:
  centroiding of profile-mode MS data} \texttt{MSnbase} vignette.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{./figure/centroiding.pdf}
  \caption{Visualisation of data smoothing and M/Z refinement using
    \texttt{MSnbase}. (a) Raw MS profile data for serine. Upper panel
    shows the base peak chromatogram (BPC), lower panel the individual
    signals in the retention time -- M/Z space. The horizontal dashed
    red line indicates the theoretical M/Z of the [M+H]+ adduct of
    serine. (b) Smoothed and centroided data for serine with M/Z
    refinement. The horizontal red dashed line indicates the
    theoretical M/Z for the [M+H]+ ion and the vertical red dotted
    line the position of the maximum signal. }
  \label{fig:cent}
\end{figure}

@@

The first public commit to the \texttt{MSnbase} GitHub repository was
in October 2010. Since then, the package benefited from 12
contributors\cite{contribs} that added various features, some
particularly significant ones such as the on-disk backend described
above. According to the package's Bioconductor page, there are 36
Bioconductor packages that depend, import or suggest it. Among these
are \texttt{pRoloc}~\cite{Gatto:2014a}, \texttt{DAPAR} and
\texttt{ProStaR}~\cite{Wieczorek:2017},
\texttt{RMassBank}~\cite{Stravs:2013},
\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
\texttt{MSstatsQC}~\cite{Dogu:2017} and the widerly used
\texttt{xcms}~\cite{Smith:2006}. \texttt{MSnbase} is also used in
packages that are not on Bioconductor, such as
IsoProt~\cite{Griss:2019}, for examples. The package currently ranks
101 out of 1823 packages based on the monthly downloads from unique IP
addresses, tallying over 1000 downloads from unique IP addresses each
months.

As is custom with Bioconductor packages, \texttt{MSnbase} comes with
ample documentation. Every user-accessible function is documented in a
dedicated manual page. In addition, the package offers 5 vignettes,
including one aimed at developers. Question from users and developers
are answered on the Bioconductor support forum as well as on the
package GitHub page. The package provides several sample and
benchmarking datasets, and relies on other dedicated
\textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
data.

\texttt{MSnbase} is available on Windows, Mac OS and Linux under the
open source Artistic 2.0 license and easily installable using standard
installation procedures.

Future developments.

@@

The version of \texttt{MSnbase} used in this manuscritp is version
\Sexpr{v}. The main features presented here were available since
version 2.0.

>>>>

@@

\title[MSnbase version 2]
  {\texttt{MSnbase}, efficient R-based processing and visualisation of raw mass spectrometry data}

@@

Mass spectrometry is a powerful technology to assays chemical and
biological samples. It is used routinely with well characterised
protocol, as well a development platform, to improve on existing
methods and devise new ones. The complexity and diversity of mass
spectrometry yields data that is itself complex and often times of
considerable size, that requires non trivial processing before
producing interpretable results. This is particularly relevant, and
can constitute a significant challenge for method developers: in
addition to the development of sample processing and mass spectrometry
methods, there is a need to process, analyse, interpret and assess
these new data to demonstrate the improvement in their technical and
analytical workflows.

There exists a very diverse catalogue of software tools to explore,
process and interpret mass spectrometry data. These range from low
level software libraries that are aimed at programmers to develop new
applications, to user-oriented applications. Example of software
libraries include Java-based jmzML~\cite{Cote:2010} or C/C++-based
ProteoWizard~\cite{Chambers:2012}, that are used as application
programming interfaces. ProteomeDiscoverer (Thermo Scientific),
MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are
among the most widely user-centric application. They generally provide
a limited and fixed set of functionality featuring often advanced
graphical user interfaces (GUI).

In this software note, we present version 2 of the \texttt{MSnbase}
\cite{Gatto:2012} software package, available from the
Bioconductor~\cite{Huber:2015} project since 2011. \texttt{MSnbase},
like other Python-based software packages such as
{pyOpenMS}~\cite{Rost:2014}, spectrum\_utils~\cite{Bittremieux:2020}
or Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies
between low level libraries and end-use software. \texttt{MSnbase}
provides a flexible R~\cite{R} command line environment for
metabolomics and proteomics mass spectrometry-based application. It
allows a detailed step-by-step processing, analysis and exploration of
the data and development of novel computational mass spectrometry
methods.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a table with
variables along the columns and each row associated to a
spectrum. Among the metadata available for each spectrum, there are MS
level, acquisition number, retention time, precursor m/z and intensity
(for MS level 2 and above), and many more. \texttt{MSnbase} provides a
rich interface to manipulate such objects. The code chunk below
illustrates such an object as displayed in the R console and an
enumeration of the metadata fields.

@@

In the following sections, we will describe demonstrate how
\texttt{MSnbase} effichiently manages large raw mass spectrometry data
and how it is used for data processing and visualisation. We will also
illustrate how it makes use of the forward-pipe operator
(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable pipelines.

@@

\item Identify and filter the groups of spectra that represent
  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)).. This can be
  done using the `filterString` metadata variable that identifies
  BoxCar spectra by their adjacent m/z segments with the
  \texttt{bc\_groups()} function and filtering relevant spectra with
  the \texttt{filterBoxCar()}.

@@

The R environment is well known for the quality of its visualiation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data for the serine and the same data after
smoothing, centroiding and m/z refinement. Detailed execution and
description of these operations can be found in the \emph{MSnbase:
  centroiding of profile-mode MS data} \texttt{MSnbase} vignette.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{./figure/centroiding.pdf}
  \caption{Visualisation of data smoothing and m/z refinement using
    \texttt{MSnbase}. (a) Raw MS profile data for serine. Upper panel
    shows the base peak chromatogram (BPC), lower panel the individual
    signals in the retention time -- m/z space. The horizontal dashed
    red line indicates the theoretical m/z of the [M+H]+ adduct of
    serine. (b) Smoothed and centroided data for serine with m/z
    refinement. The horizontal red dashed line indicates the
    theoretical m/z for the [M+H]+ ion and the vertical red dotted
    line the position of the maximum signal. }
  \label{fig:cent}
\end{figure}

@@

The first public commit to the \texttt{MSnbase} GitHub repository was
in October 2010. Since then, the package benefited from 12
contributors\cite{contribs} that added various features, some
particularly significant ones such as the on-disk backend described
above. According to the package's Bioconductor page, there are 36
Bioconductor packages that depend, import or suggest it. Among these
are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
spectrometry-based spatial proteomics data,
\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
\texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
statistical analysis for quantitative proteomics data,
\texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
MS files and build MassBank records,
\texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system
suitability monitoring and quality control of targeted proteomic
experiments and the widely used \texttt{xcms}~\cite{Smith:2006} packge
for the processing and analysis of metabolomics data. \texttt{MSnbase}
is also used in non-R/Bioconductor software, such as for example
IsoProt~\cite{Griss:2019}, that provides a reproducible workflow for
iTRAQ/TMT experiments. \texttt{MSnbase} currently ranks 101 out of
1823 packages based on the monthly downloads from unique IP addresses,
tallying over 1000 downloads from unique IP addresses each months.

As is custom with Bioconductor packages, \texttt{MSnbase} comes with
ample documentation. Every user-accessible function is documented in a
dedicated manual page. In addition, the package offers 5 vignettes,
including one aimed at developers. Question from users and developers
are answered on the Bioconductor support forum as well as on the
package GitHub page. The package provides several sample and
benchmarking datasets, and relies on other dedicated
\textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
data. \texttt{MSnbase} is available on Windows, Mac OS and Linux under
the open source Artistic 2.0 license and easily installable using
standard installation procedures.

@@

The version of \texttt{MSnbase} used in this manuscript is
\Sexpr{v}. The main features presented here were available since
version 2.0.

### 2020-04-27 msnbase2.tex
first draft ready

The first public commit to the \texttt{MSnbase} GitHub repository was
in October 2010. Since then, the package benefited from 12
contributors\cite{contribs} that added various features, some
particularly significant ones such as the on-disk backend described
above. According to the package's Bioconductor page, there are 36
Bioconductor packages that depend, import or suggest it. Among these
are \texttt{pRoloc}~\cite{Gatto:2014a}, \texttt{DAPAR} and
\texttt{ProStaR}~\cite{Wieczorek:2017},
\texttt{RMassBank}~\cite{Stravs:2013},
\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
\texttt{MSstatsQC}~\cite{Dogu:2017} and the widerly used
\texttt{xcms}~\cite{Smith:2006}. \texttt{MSnbase} is also used in
packages that are not on Bioconductor, such as
IsoProt~\cite{Griss:2019}, for examples. The package currently ranks
101 out of 1823 packages based on the monthly downloads from unique IP
addresses, tallying over 1000 downloads from unique IP addresses each
months.

As is custom with Bioconductor packages, \texttt{MSnbase} comes with
ample documentation. Every user-accessible function is documented in a
dedicated manual page. In addition, the package offers 5 vignettes,
including one aimed at developers. Question from users and developers
are answered on the Bioconductor support forum as well as on the
package GitHub page. The package provides several sample and
benchmarking datasets, and relies on other dedicated
\textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
data.

\texttt{MSnbase} is available on Windows, Mac OS and Linux under the
open source Artistic 2.0 license and easily installable using standard
installation procedures.

Future developments.

The version of \texttt{MSnbase} used in this manuscritp is version
2.13.9. The main features presented here were available since
version 2.0.

>>>>

The first public commit to the \texttt{MSnbase} GitHub repository was
in October 2010. Since then, the package benefited from 12
contributors\cite{contribs} that added various features, some
particularly significant ones such as the on-disk backend described
above. According to the package's Bioconductor page, there are 36
Bioconductor packages that depend, import or suggest it. Among these
are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
spectrometry-based spatial proteomics data,
\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
\texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
statistical analysis for quantitative proteomics data,
\texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
MS files and build MassBank records,
\texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system
suitability monitoring and quality control of targeted proteomic
experiments and the widely used \texttt{xcms}~\cite{Smith:2006} packge
for the processing and analysis of metabolomics data. \texttt{MSnbase}
is also used in non-R/Bioconductor software, such as for example
IsoProt~\cite{Griss:2019}, that provides a reproducible workflow for
iTRAQ/TMT experiments. \texttt{MSnbase} currently ranks 101 out of
1823 packages based on the monthly downloads from unique IP addresses,
tallying over 1000 downloads from unique IP addresses each months.

As is custom with Bioconductor packages, \texttt{MSnbase} comes with
ample documentation. Every user-accessible function is documented in a
dedicated manual page. In addition, the package offers 5 vignettes,
including one aimed at developers. Question from users and developers
are answered on the Bioconductor support forum as well as on the
package GitHub page. The package provides several sample and
benchmarking datasets, and relies on other dedicated
\textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
data. \texttt{MSnbase} is available on Windows, Mac OS and Linux under
the open source Artistic 2.0 license and easily installable using
standard installation procedures.

The version of \texttt{MSnbase} used in this manuscript is
2.13.9. The main features presented here were available since
version 2.0.
